Addressing selection bias and measurement error in COVID-19 case count data using auxiliary information

### 2022-03-28 write-up/covid-supplement-writeup.tex
fixed new exp to new inf and now all good

For simplicity, we list the priors used in STAN below:
\begin{itemize}
  \item \code{beta $\sim$ Normal(1.5, 1) T[0,];}
  \item \code{gamma $\sim$ normal(0.3, 0.5) T[0,];}
  \item \code{sigma $\sim$ normal(0.4, 0.5) T[0,];}
  \item \code{phi\_inv $\sim$ exponential(5);}
  \item \code{i0 $\sim$ normal(0, 10);}
  \item \code{e0 $\sim$ normal(0, 10);}
  \item \code{eta $\sim$ normal(1, 1) T[0,];}
  \item \code{eta\_two $\sim$ normal(1, 1) T[0,];}
  \item \code{eta\_three $\sim$ normal(1, 1) T[0,];}
  \item \code{nu $\sim$ exponential(1./5);}
  \item \code{nu\_two $\sim$ exponential(1./5);}
  \item \code{nu\_three $\sim$ exponential(1./5);}
  \item \code{xi\_raw $\sim$ beta(1, 1);}
  \item \code{phi = 1/phi\_inv;}
  \item \code{xi = xi\_raw + 0.5;}
\end{itemize}
where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}. Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.  \textcolor{red}{Sub strata via kernel weighting due to limited data.}

Figure~\ref{fig:undercounting} below is a plot of undercount based on the SEIR model.  We see the undercount goes away by January; however, this points to potential misspecification of the IFR.

@@

Figure~\ref{fig:tv_air_sens} below is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.  We see that the impact on the doubly robust estimates is minimal.

>>>>

For simplicity, we list the priors used in STAN below:
\begin{itemize}
  \item \code{beta $\sim$ Normal(1.5, 1) T[0,];}
  \item \code{gamma $\sim$ normal(0.3, 0.5) T[0,];}
  \item \code{sigma $\sim$ normal(0.4, 0.5) T[0,];}
  \item \code{phi\_inv $\sim$ exponential(5);}
  \item \code{i0 $\sim$ normal(0, 10);}
  \item \code{e0 $\sim$ normal(0, 10);}
  \item \code{eta $\sim$ normal(1, 1) T[0,];}
  \item \code{eta\_two $\sim$ normal(1, 1) T[0,];}
  \item \code{eta\_three $\sim$ normal(1, 1) T[0,];}
  \item \code{nu $\sim$ exponential(1./5);}
  \item \code{nu\_two $\sim$ exponential(1./5);}
  \item \code{nu\_three $\sim$ exponential(1./5);}
  \item \code{xi\_raw $\sim$ beta(1, 1);}
  \item \code{phi = 1/phi\_inv;}
  \item \code{xi = xi\_raw + 0.5;}
\end{itemize}
where \code{beta, eta, eta\_two, eta\_three} refer to the four values that form the time-varying parameter $\beta_t$ when combined with \code{xi, eta, eta\_two, eta\_three} in the SEIR model as described in Section~\ref{section:modelbased}. Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.  Due to the low number of deaths per strata, i.e., based on Race, Age, Sex and Ethnicity, we employ a kernel smoothing of the new infections per age strata into these sub-strata using relative number of observed deaths.  This ensured a suitably complex model that could fit the data while generating strata-specific active infection rates that can be used in the doubly-robust estimation procedure.

@@

Figure~\ref{fig:undercounting} is a plot of undercount based on the SEIR model.  We plot relative undercound on a cumulative basis since January due to the case counts reflecting active infections. Figure~\ref{fig:tv_air_sens} is a sensitivity analysis of the active infection rate under SEIR model with 10\% increase and decrease in the average IFR.  We see that the impact on the doubly robust estimates is minimal.

### 2022-03-28 write-up/covid-writeup.tex
fixed new exp to new inf and now all good

 Up to this point, the primary focus has been selection bias in coronavirus case-count data from a survey sampling perspective.  Here, we consider compartmental model approaches from infectious disease epidemiology.  Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$. To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented. A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.

 Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.  The population-level SEIR dynamics are given by the set of differential equations in~\eqref{eq:seir}.
% \begin{align}
%  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
%  \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
%  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
% \end{align}
%Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.  Let $I_{k,t}$ to denote the number of new infections on day $t$ in the $k$th strata.  The joint distribution of strata-specific new infection totals follows a multinomial distribution with
 $$
 ( I_{1,t}, \ldots, I_{K,t} ) \sim \text{Multinomial} \left( I_t, (p_{1,t}, \ldots, p_{K, t}) \right).
 $$
 where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.

 Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
 $$
 D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I_{k,t} \theta_{(r-t)} \right)
 $$
 where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}. See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.  %Additional details on inference are discussed in Appendix~\ref{app:modelbased}.

@@

 \begin{remark}[Sensitivity to IFR]

 \end{remark}

  \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
 \caption{PPC on aggregate daily death counts.}
 \label{fig:deathsppc}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/cases_ppc.png}
 \caption{Posterior on daily case counts.}
 \label{fig:casesppc}
 \end{subfigure}
 \caption{Posterior predictive check (PPC) on death counts and posterior distribution of case counts.}
 \label{fig:ppcs}
 \end{figure}

@@

 To check model fit, a posterior predictive check on aggregate death counts was performed. Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well. Figure~\ref{fig:casesppc} presents the posterior distribution of total new cases.  Figure~\ref{fig:undercounting} in the supplementary materials computes the weekly undercount factor of observed case counts.  Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.

>>>>

 Up to this point, the primary focus has been selection bias in coronavirus case-count data from a survey sampling perspective.  Here, we consider compartmental model approaches from infectious disease epidemiology.  Our primary objective is a model-based forecast of strata-level active infection rate~$\bar y_k$. To do this, a probabilistic extension of a standard epidemiological state-space model -- the susceptible, exposed, infected, and removed (recovered and death) model, or SEIR model -- is presented. A probabilistic SIR model was originally proposed by~\cite{Osthus2017} with only one-dimensional time series of infected proportions; this formulation was extended by~\cite{Song2020} to model coronavirus case-counts.

 Let $s_t$, $e_t$, $i_t$, and $r_t$ denote the proportion of survivors, exposed, infected, and removed cases (i.e., including both recovered cases and deaths) at time $t$.  The population-level SEIR dynamics are given by the set of differential equations in~\eqref{eq:seir}.
% \begin{align}
%  \frac{\partial s_t}{\partial t} &= - \beta s_t i_t; \quad \frac{\partial e_t}{\partial t} = \beta s_t i_t - \sigma e_t; \quad \label{eq:seir} \\
%  \frac{\partial i_t}{\partial t} &= \sigma e_t - \gamma i_t; \quad
%  \frac{\partial r_t}{\partial t} = - \gamma i_t. \nonumber
% \end{align}
%Following~\cite{Song2020}, we employ Runge-Kutta (RK4) approximations for discretization.
Here, we consider covariate information that takes the form of a stratification variable with $K$ strata.  Based on these population-level dynamics, we can compute the number of \emph{new infections} at time~$t$, i.e., $I^\new_{t} := - N \cdot (e_{t+1} - e_{t} + s_{t+1} - s_t)$ to denote the number of new exposures on day $t$ in the $k$th strata.  The joint distribution of strata-specific new exposures~$ \{ I^{\new}_{t,k} \}_{k=1}^K$ follows a multinomial distribution with
 $$
 ( I^\new_{1,t}, \ldots, I^\new_{K,t} ) \sim \text{Multinomial} \left( I^\new_t, (p_{1,t}, \ldots, p_{K, t}) \right).
 $$
 where $( {\bf p}_t )_{t=1}^T = ( (p_{1,t}, \ldots, p_{K, t}) )_{t=1}^T$ are a sequence of parameters on the simplex, i.e., $\sum_{k=1}^K p_{k,t} = 1$.

 Selection bias is addressed by analyzing COVID-19 death data rather than case counts.  Let $D_{k,r}$ denote the number of individuals who pass away on day $r$ from strata $k$.  Then
 $$
 D_{k,r} \mid p, \theta, \nu \sim \text{Poisson} \left( \sum_{t=1}^r \text{IFR}_{k} \cdot I^\new_{k,t} \theta_{(r-t)} \right)
 $$
 where $\text{IFR}_{k}$ is the infection fatality rate for the $k$th strata and $\theta_{j}$ is a discrete-time distribution for time from infection to death.  This extends prior analysis of death data~\cite{Johndrow2020} by allowing the infection fatality rate to depend on a stratification variable, which is important as IFR depends heavily on age~\citep{Levin2020}. See Section~\ref{section:application_of_modelbased} for discussion of parameter choices.  %Additional details on inference are discussed in Appendix~\ref{app:modelbased}.

@@

 \begin{remark}[Sensitivity to IFR]
To account for potential misspecification of the marginal IFR, we perform the same analysis with a 10\% increase and decrease in the marginal IFR under the log-linear model. Figure~\ref{fig:tv_air_sens} in the supplementary materials presents the model-based and doubly robust estimates in these scenarios.
\end{remark}

  \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
 \caption{PPC on aggregate daily death counts.}
 \label{fig:deathsppc}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/infections_ppc.png}
 \caption{Posterior on daily new infections.}
 \label{fig:casesppc}
 \end{subfigure}
 \caption{Posterior predictive check (PPC) on death counts and posterior distribution of infections.}
 \label{fig:ppcs}
 \end{figure}

@@

 To check model fit, a posterior predictive check on aggregate death counts was performed. Figure~\ref{fig:deathsppc} shows the results which suggests chosen SEIR model fits the observed data well. Figure~\ref{fig:casesppc} presents the posterior distribution of total new infections. As the active infection rate is defined as all infected individuals who’s viral load has yet to fall below detectable levels by RT-PCR testing, to construct the active infection estimates we exponentially discount the number of newly infected individuals to estimate the number of individuals with an active infection on each day. In our analysis, the exponential discounting is set to $20$ days to match with prior evidence that ``30\% to 40\% of people will still test positive at three weeks''~\citep{wlocneg}. To compare with case count data, Figure~\ref{fig:undercounting} in the supplementary materials computes the cumulative undercount factor of observed case counts as compared to new infections.  Posterior distributions for the age-specific case counts are used as part of the doubly-robust estimator of the active infection rate.
